---
title: "EV Stations Analysis - State Level"
author: "Andy Pickering"
format: html
editor: visual
toc: true
code-link: true
---

# Analyzing US EV charging stations by State

# Load data and libraries

Data is downloaded and saved in *get_ev_stations_data_US.R*

```{r}

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tigris))
options(tigris_use_cache = TRUE)
library(leaflet)
suppressPackageStartupMessages(library(rmapshaper))
library(janitor)

data_date <- "2023-08-08"

response <- readRDS(paste0('./data/',data_date,'_Elec_Stations_US_All'))
#ev <- readRDS(paste0('./data/',data_date,'_Elec_Stations_US_All_df'))
ev <- readRDS(paste0('./data/',data_date,'_Elec_Stations_US_All_df_levels_connectors'))

```

# Analysis

## Count total number of stations by state

```{r count by state}

state_counts <- 
  ev %>% count(state) %>% 
  arrange(desc(n)) %>% 
  as_tibble()

#View(state_counts)
head(state_counts,20)
```

```{r bar chart stations per state}

state_counts %>%
  mutate(state = forcats::fct_reorder(state, n)) %>% 
  dplyr::slice_max(`n`, n = 20) %>% 
  ggplot(aes(state, n)) +
  geom_col() +
  ggtitle("Number of EV Charging Stations Per State (top 20)") +
  xlab("State") +
  ylab("Number of EV Stations") +
  coord_flip()


```

## chloropleth map with number of stations by state

### ggplot

```{r get states shapefiles}

states_map <- tigris::states() %>% 
  janitor::clean_names() %>% 
  mutate(statefp = as.numeric(statefp)) %>% 
  filter(statefp < 60,
         !statefp %in% c(2, 15)) %>% 
  rmapshaper::ms_simplify()


```

Join ev counts data to states shapefile

```{r join data}

dat_to_map <- states_map %>% 
  left_join(state_counts, by = c("stusps" = "state"))

```

Make choropleth using ggplot2

```{r ggplot chorpleth}

ggplot() +
  geom_sf(data = dat_to_map,
          aes(fill = n)) +
  scale_fill_viridis_c() +
  xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('Number of EV Charging Stations Per State') +
  theme_void()

  
```

### leaflet chorpleth

```{r leaflet choropleth}

pal_ev <- leaflet::colorNumeric(palette = "viridis",
                                 domain = dat_to_map$n)

states_ev_map <- leaflet() %>% 
#  addTiles() %>% # adds OpenStretMap basemap
  addPolygons(data = dat_to_map,
              weight = 1,
              color = "black",
              popup = paste(dat_to_map$name, "<br>",
                            " EV Stations: ", dat_to_map$n, "<br>"),
              fillColor = ~pal_ev(n),
              fillOpacity = 0.6) %>% 
  addLegend(data = dat_to_map,
            pal = pal_ev,
            values = ~n,
            opacity = 1,
            title = "# EV Stations <br>
            Per State"
            )


states_ev_map

```

## Break down data in more detail

-   public vs. private, charging level etc.

### Make a function to plot choropleth for any variable

```{r}

make_leaflet_choropleth <- function(dat_to_map, data_title){

pal_ev <- leaflet::colorNumeric(palette = "viridis",
                                 domain = dat_to_map$n)

the_map <- leaflet() %>% 
#  addTiles() %>% # adds OpenStretMap basemap
  addPolygons(data = dat_to_map,
              weight = 1,
              color = "black",
              popup = paste(dat_to_map$name, "<br>",
                            " EV Stations: ", dat_to_map$n, "<br>"),
              fillColor = ~pal_ev(n),
              fillOpacity = 0.6) %>% 
  addLegend(data = dat_to_map,
            pal = pal_ev,
            values = ~n,
            opacity = 1,
            title = paste("# EV Stations <br>",
                          data_title)
            )

return(the_map)

}
```

### Public Stations Only

```{r}

state_counts_public <- 
  ev %>% 
  filter(access_code == "public") %>% 
  count(state) %>% 
  arrange(desc(n))

dat_to_map <- states_map %>% 
  left_join(state_counts_public, by = c("stusps" = "state"))

the_map <- make_leaflet_choropleth(dat_to_map, data_title = "public")

the_map

```

### Private Stations

```{r}

state_counts_private <- 
  ev %>% 
  filter(access_code == "private") %>% 
  count(state) %>% 
  arrange(desc(n))

dat_to_map <- states_map %>% 
  left_join(state_counts_private, by = c("stusps" = "state"))

the_map <- make_leaflet_choropleth(dat_to_map, data_title = "private")

the_map

```

### Breakdown by charger level/type

#### Count by level2 / level3 (# stations for now, not \# chargers)

```{r}

state_counts_by_level <- ev %>% 
  group_by(state) %>% 
  summarize(nL1 = sum(has_L1, na.rm = TRUE),
            nL2 = sum(has_L2, na.rm = TRUE),
            nDC = sum(has_DC, na.rm = TRUE)
            )

state_counts_L2 <- ev %>% 
  filter(has_L2 == TRUE) %>% 
  group_by(state) %>% 
  count(has_L2) %>% 
  arrange(desc(n))


state_counts_DC <- ev %>% 
  filter(has_DC == TRUE) %>% 
  group_by(state) %>% 
  count(has_DC) %>% 
  arrange(desc(n))

```

```{r}


dat_to_map <- states_map %>% 
  left_join(state_counts_DC, by = c("stusps" = "state"))

make_leaflet_choropleth(dat_to_map, data_title = "DC")

```

```{r}


dat_to_map <- states_map %>% 
  left_join(state_counts_L2, by = c("stusps" = "state"))

make_leaflet_choropleth(dat_to_map, data_title = "Level2")

```
