---
title: "EV Charging Stations Analysis"
author: "Andy Pickering"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, echo=FALSE, message=FALSE}

library(httr)
library(jsonlite)
library(ggplot2)
library(dplyr)

```

## Load Data For EV stations in Colorado

- I'll retrieve the EV station data using the AFDC API. The documentation for the AFDC fuel-stations API can be found at: <https://developer.nrel.gov/docs/transportation/alt-fuel-stations-v1/all/#station-count-record-fields>

- You can obtain a free API key at: [<https://developer.nrel.gov/signup/>]. I've saved my API key in my local .Renviron file so I can load it without exposing the key in my code.

- For now I will request all EV (electric fuel type) stations in Colorado.

```{r Load Data from API}

# API key is stored in my .Renviron file
api_key <- Sys.getenv("AFDC_KEY")

# base url for AFDC alternative fuel stations API
target <- "https://developer.nrel.gov/api/alt-fuel-stations/v1"

# Return data for all electric stations in Colorado
api_path <-".json?&fuel_type=ELEC&state=CO&limit=all"

complete_api_path <- paste0(target,api_path,'&api_key=',api_key)

dat <- httr::GET(url=complete_api_path)

dat2 <- jsonlite::fromJSON(httr::content(dat,"text"))

print(paste(dat2$total_results,'Stations Found'))

df <- dat2$fuel_stations

#str(df)

```

### Filter out non-EV data columns

The returned data contains many non-electric fields that we don't need (they will all be NA since we requested electric fuel type only):

```{r }

colnames(df)

```

I'll remove a bunch of the non-relevant fields from the data frame to clean things up a bit. I'll also change the date column type and add a variable for year opened.

```{r remove fields}

# filter out non-EV related fields
df <- df %>% select(-starts_with("lng")) %>% 
  select(-starts_with("cng")) %>%
  select(-starts_with("lpg")) %>%
  select(-starts_with("hy")) %>% 
  select(-starts_with("ng")) %>% 
  select(-starts_with("e85")) %>% 
  select(-starts_with("bd")) %>% 
  select(-starts_with("rd"))


# change date field to date type and add a year opened variable
df$open_date <- lubridate::ymd(df$open_date)
df$open_year <- lubridate::year(df$open_date)

colnames(df)
#str(df)

```


## Station Openings Over Time

### Look at how many stations opened each year

First I'd like to look at how many EV stations opened over time, so I'll make a new data frame summarizing the number of stations opened by year.

```{r }
df_opened <- df %>% 
  count(open_year,name = "nopened")  

df_opened
```

#### Plot Number of Stations Opened Each year

```{r }
df_opened %>% ggplot(aes(open_year, nopened)) + 
  geom_col()+
  xlab("Year Opened")+
  ylab("# Stations Opened")+
  ggtitle('EV Stations Opened in Colorado Each Year')

```

#### Plot cumulative sum of stations opened over time

We can also look at the cumulative sum of stations opened over time

```{r}

df_opened %>% ggplot(aes(open_year,cumsum(nopened)))+
  geom_line()+
  geom_point()+
  xlab("Year")+
  ylab("# Stations")+
  ggtitle("Cumulative sum of EV stations opened in CO")

```

### Station openings by level/charger type

Next I want to dig a little deeper and break down the station openings by charger type and/or level. I'd expect to see an increase in DC fast charging stations in more recent years. I'll make a new data frame with the number of chargers opened by year, grouped by charging level (Level 1, Level 2, or DC fast).

```{r}

df_opened_level <- df %>% select(id,open_date,open_year,ev_dc_fast_num,ev_level2_evse_num,ev_level1_evse_num) %>%
  group_by(open_year) %>%
  summarize(n_DC=sum(ev_dc_fast_num,na.rm = TRUE), 
            n_L2=sum(ev_level2_evse_num,na.rm = TRUE),
            n_L1=sum(ev_level1_evse_num,na.rm = TRUE) )

df_opened_level

```

To make plotting easier, I'll pivot ('melt') the dataframe from wide to long format so I can group by charnging level:

```{r}

df_opened_level_long <- df_opened_level %>% 
  tidyr::pivot_longer(cols=c('n_DC','n_L2','n_L1'), names_to = "Level", names_prefix = "n_", values_to = "n_opened")

df_opened_level_long


```


```{r }

df_opened_level_long %>% 
  ggplot(aes(open_year, n_opened, group=Level)) +
  geom_line(aes(col=Level))+
  geom_point(aes(col=Level))+
  xlab("Year Opened")+
  ylab("# Charges Opened")+
  ggtitle("Number of Chargers Opened Per Year")
  

```

Cumulative sum of chargers added by level

```{r}

df_opened_level %>% ggplot()+
  geom_line(aes(open_year,cumsum(n_DC)),col='black')+
  geom_line(aes(open_year,cumsum(n_L2)),col='red')+
  geom_line(aes(open_year,cumsum(n_L1)),col='green')

```



## Count \# individual plugs by level

```{r }

n_lev1 <- sum(df$ev_level1_evse_num,na.rm=TRUE)
n_lev2 <- sum(df$ev_level2_evse_num,na.rm=TRUE)
n_dcfast <- sum(df$ev_dc_fast_num, na.rm=TRUE)

plugs <- tibble(level=c('Level1','Level2','DC Fast'), n=c(n_lev1,n_lev2,n_dcfast))

plugs %>% ggplot(aes(level,n))+
  geom_col()+
  ggtitle("Total Number of Plugs By Charging Level")

```

<!-- ## Break down by connector types, ev_connector_type,ev_charging_level -->

<!-- ```{r } -->

<!-- #unique(unlist(df$ev_connector_types)) -->

<!-- df4 <- df3 %>% rowwise() %>%  -->

<!--   mutate(hasJ=if_else("J1772" %in% ev_connector_types,1,0)) %>%  -->

<!--   mutate(hasJcomb=if_else("J1772COMBO" %in% ev_connector_types,1,0)) %>%  -->

<!--   mutate(hasChad=if_else("CHADEMO" %in% ev_connector_types,1,0)) %>%  -->

<!--   mutate(hasTesla=if_else("TESLA" %in% ev_connector_types,1,0)) %>%  -->

<!--   mutate(hasNema520=if_else("NEMA520" %in% ev_connector_types,1,0)) %>%  -->

<!--   mutate(hasNema515=if_else("NEMA515" %in% ev_connector_types,1,0)) %>%  -->

<!--   mutate(hasNema1450=if_else("NEMA1450" %in% ev_connector_types,1,0)) -->

<!--   View(df4) -->

<!-- df4 %>% -->

<!--   select_if(is.numeric) %>% -->

<!--   map_dbl(sum)   -->

<!-- tibble(connector_type=names(df5),n=df5) %>% View() -->

<!-- ``` -->

<!-- ```{r } -->

<!-- #df %>% count(ev_network) %>% arrange(desc(n)) %>% View() -->

<!-- df %>%  count(ev_network) %>%  -->

<!--   mutate(ev_network=as.factor(ev_network)) %>%  -->

<!--   mutate(ev_network=forcats::fct_reorder(ev_network,n)) %>%  -->

<!--   ggplot(aes(ev_network,n))+ -->

<!--   geom_col()+ -->

<!--   coord_flip()+ -->

<!--   xlab("EV Network")+ -->

<!--   ylab("# stations") -->

<!-- ``` -->

## Break down by city/county

```{r}

df %>% count(city) %>%
  slice_max(order_by=n,n=15) %>%
  mutate(city=forcats::fct_reorder(city,n)) %>%
  ggplot(aes(city,n))+
  geom_col()+
  coord_flip()+
  ggtitle("Number of EV Stations by City (Top 15)")+
  xlab("City")+
  ylab("Number of stations")

#  View()

```
